---
title: "Assignment3"
output: pdf_document
date: "2025-02-13"
---

```{r setup, include=FALSE}
library(ISLR)
data(Hitters)
Hitters.new <- na.omit(Hitters)
X <- as.matrix(Hitters.new[,c(-14,-15,-20,-19)])
Y <- as.matrix(Hitters.new[,19])

```

Using mallow's cp
```{r}
library(leaps)

subset_model <- regsubsets(Y ~ ., data = Hitters.new, nvmax = ncol(X))

subset_summary <- summary(subset_model)

best_model_cp <- which.min(subset_summary$cp)

best_variables_cp <- names(coef(subset_model, best_model_cp))
print(best_variables_cp)

```

Adjusted R² and BIC:
```{r}
best_model_r2 <- which.max(subset_summary$adjr2)
best_variables_r2 <- names(coef(subset_model, best_model_r2))

best_model_bic <- which.min(subset_summary$bic)
best_variables_bic <- names(coef(subset_model, best_model_bic))

print(best_variables_r2)  
print(best_variables_bic) 

```

using step:
```{r}
full_model <- lm(Salary ~ ., data = Hitters.new)

backward_model <- step(full_model, direction = "backward", trace = FALSE)

null_model <- lm(Salary ~ 1, data = Hitters.new)

forward_model <- step(null_model, scope = formula(full_model), direction = "forward", trace = FALSE)


stepwise_model <- step(null_model, scope = formula(full_model), direction = "both", trace = FALSE)

summary(backward_model)
summary(forward_model)
summary(stepwise_model)

```

QUESTION 2
a)
```{r}
data(iris)

iris_data <- iris[, 1:4]

cov_matrix <- var(iris_data)

eigen_decomp <- eigen(cov_matrix)

print(eigen_decomp$values)
print(eigen_decomp$vectors)

pca_result <- princomp(iris_data, cor = FALSE)
summary(pca_result)  
print(pca_result$loadings)

```

b)
```{r}

pc1 <- eigen_decomp$vectors[, 1]
pc2 <- eigen_decomp$vectors[, 2]

print(pc1)  
print(pc2)  

```

c)
Argument for:
In case variables have different units, variable differences can dominate PCA.

Argument against: 
If raw measurements already have meaning. PCA can distort them.

```{r}
pca_scaled <- princomp(iris_data, cor = TRUE)
summary(pca_scaled)

```

d)
```{r}
pc_projection <- as.matrix(iris_data) %*% eigen_decomp$vectors[, 1:2]


iris_pca_df <- data.frame(PC1 = pc_projection[, 1], PC2 = pc_projection[, 2], Species = iris$Species)

library(ggplot2)

ggplot(iris_pca_df, aes(x = PC1, y = PC2, color = Species)) +
  geom_point(size = 3) +
  labs(title = "PCA Projection of Iris Data", x = "PC1", y = "PC2") +
  theme_minimal()

```

e)
```{r}
perform_pca <- function(species_name) {
  species_data <- iris_data[iris$Species == species_name, ]
  pca_result <- princomp(species_data, cor = FALSE)
  return(summary(pca_result))
}

pca_setosa <- perform_pca("setosa")
pca_versicolor <- perform_pca("versicolor")
pca_virginica <- perform_pca("virginica")

print(pca_setosa)
print(pca_versicolor)
print(pca_virginica)

```
The first principal component in each species often represents variation in flower size.
The differences suggest species have different sources of variance.
When PCA is performed on the full dataset, interspecies variation influences the results, whereas within-species PCA highlights species-specific variation.
